Designing an effective intervention of motivational interview using sequence classification methods

### 2017-03-02 paper/AMIA/main.tex
update results and discussion sections

if $p(S\rightarrow CT) > \delta $, where $\delta$ is experimentally defined threshold, the interview transcript (or a portion of it) is predicted to result in CT. We experimentally determined the value of $\delta$ to 0.3 which provides acceptable accuracy with minimum false positive rate. Figure\ref{fig:delta} illustrates the characteristics of proposed model by varying the value of $\delta$.

@@

\subsection*{\textit{Performance of models for the classification of normal PPC code sequences}}
Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}. From Table~\ref{table:sequence}, it follows that our designed Markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092. The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}. It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\

\begin{table}[h]
\centering
\caption{Performance of Markov models for the classification of normal PPC code sequence.}
\label{tab:result_norm_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

@@

Since clinical researchers typically annotate caregiver and adolescent sessions separately, we first created two experimental data sets consisting of only adolescent and only caregiver session transcripts. Second, besides evaluating the accuracy of annotating adolescent and caregiver transcripts with the codebooks containing an entire set of codes, we also conducted a series of experiments with the codebooks of smaller sizes created as outlined above. Third, besides training and testing NB, SVM, CRF, Decision Tree, Boosting, DiscLDA, Random Forest and CNN classifiers using only lexical features, we also evaluated the effectiveness of the proposed contextual and semantic features. Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.

\subsection*{\textit{Performance of models for the classification of alternate PPC code sequences}}
Two Markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses. Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences. Two Markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses. Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences. Two Markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses. Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences. \\

\begin{table}[h]
\centering
\caption{Performance of Markov models for the classification of alternate PPC code sequence.}
\label{tab:result_alt_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

@@

In general, these higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the data set. However, the number of states in higher order Markov chains grows exponentially with the order of the chain. Consequently, it is harder to accurately estimate all the transition probabilities from the training set. That is, there are many high-order states that are not frequently observed in the training set, leading to inaccurate probability estimates. In principle, this problem can be solved if the size of the training set is increased at the same rate as the state-size of the higher order Markov chain. However, this cannot be easily done for many applications areas. To overcome these problems hidden Markov model is used with existing library available in online. The basic idea of these approaches is to combine portions of different order Markov chains so that to improve the overall classification accuracy.

\section*{Discussion}
From the sequential analysis, we observed that Markov model achieved near-human accuracy to categories the sequence of behavior codes. It was also found that successful interviews are more frequently responded by the adolescent with PPC code 112. However, unsuccessful interviews are responded with the behavior code 109. Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis. Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}. From Table~\ref{table:sequence}, it follows that our designed Markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092. The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}. It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.

From the sequential analysis, we observed that Markov model achieved near-human accuracy to categories the sequence of behavior codes. It was also found that successful interviews are more frequently responded by the adolescent with PPC code 112. However, unsuccessful interviews are responded with the behavior code 109. Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis. Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}. From Table~\ref{table:sequence}, it follows that our designed Markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092. The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}. It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.

>>>>

if $p(S\rightarrow CT) > \delta $, where $\delta$ is experimentally defined threshold, the interview transcript (or a portion of it) is predicted to result in CT. To fit the model, we experimentally determined the value of $\delta$ where model provides highest F-measure with lower bias. Figure\ref{fig:delta} [this is a temporary figure, I will change it later] illustrates an example of the characteristics of Markov model by varying the value of $\delta$.

@@

\subsection*{\textit{Performance of models for the classification of normal PPC code sequences}}
Performance summary for the classification of normal sequences is presented in Table~\ref{tab:result_norm_seq}. It follows that first order hidden Markov model works very well and achieves highest precision 0.7980 with F-measure 0.7989. However, our designed second order Markov model shows lowest performance in terms of precision and F-measure. It was also noticed that the lowest recall 0.7686 achieved in first order Markov model. On the other hand, second order hidden Markov model achieves the highest recall 0.8449 and accuracy 0.8449. From Table~\ref{tab:result_norm_seq}, we observed that performance degrades from first order to second order. Precision and F-measures drop 6.74\% \& 2.39\% in Markov chain and 7.27\% \& 2.09\% in HMM, respectively, when using second order models. In contrast, recall improves 2.64\%--4.84\% in second order Markov model and HMM. Overall, HMM performed better than general Markov model in normal utterance sequences.

\begin{table}[h]
\centering
\caption{Performance of Markov models for the classification of normal PPC code sequence. Highest performances for each metrics are highlighted in bold faces.}
\label{tab:result_norm_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

@@

\subsection*{\textit{Performance of models for the classification of alternate PPC code sequences}}
Table~\ref{tab:result_alt_seq} shows the classification performance of models applied on alternate sequences. The results demonstrate that second order hidden Markov model performed better than Markov chain in terms of recall and achieved 97.13\% recall. Second order general Markov model shows the highest precision 0.9778 as well as highest F-measure 0.9736. On the other hand, first order hidden Markov model displays lowest performance in all metrics compared to all other models. It was also examined that every models perform better in second order compared to first order. General Markov chain achieves 1\%, 1\% and 1\% improvement for precision, recall and F-measure, respectively, from first order to second order. Similarly, hidden Markov model achieves 1\%, 1\% and 1\% improvement for precision, recall and F-measure, respectively, from first order to second order. Overall, second order Markov model performs better for alternate utterance sequences. \\

\begin{table}[h]
\centering
\caption{Performance of Markov models for the classification of alternate PPC code sequence. Highest performances for each metrics are highlighted in bold faces.}
\label{tab:result_alt_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

@@

\section*{Discussion}
The purpose of this study was to facilitate the design of motivational interview by examining the performance of sequence classification in the field of health informatics. Studying the results across the different classification schemes we can make some key observations.

First, higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the data set. However, the number of states in higher order Markov chains grows exponentially with the order of the chain. Consequently, it is harder to accurately estimate all the transition probabilities from the training set. That is, there are many high-order states that are not frequently observed in the training set, leading to inaccurate probability estimates. In principle, this problem can be solved if the size of the training set is increased at the same rate as the state-size of the higher order Markov chain. However, this cannot be easily done for many applications areas include our case. To skip these problems we restricted our study upto second order. Shorter average length of utterance sequence is another reason for keeping the order lower.

Second, overall performance of hidden Markov model is better than general Markov model in both type of sequences. We observed that hidden Markov model achieved near-human accuracy to categories the sequences of behavior codes. This means that hidden states capture the propagated information passes through the sequence for determining final outcome.

Third, alternate sequences provide higher performance for all Markov models. Even, it provides highest precision and F-measure for alternate sequences. This indicates that each pair of alternate behavior codes forward their mutual information to the next step more accurately which improve performance for distinguishing outcomes of the motivational interview. And this is obvious because if we know the previous two behavior codes (e.g. provider's question and adolescent's answer) then it is easier to know the expected answer for the current question.

Finally, it was seen that successful interviews are more frequently responded by the adolescent with positive behavior changes which lead to make a positive commitment for loosing weight. Whereas, unsuccessful interviews are responded with negative behavior changes which lead to a negative commitment or no commitment.
